---
title: "busco2aster Visual Report"
format:
  html:
    toc: true
    toc-depth: 2
    theme: flatly
    code-fold: false
    number-sections: false
params:
  data_dir: "results/report/data"
---

```{r}
#| label: setup
#| include: false
library(ape)
library(dplyr)
library(ggplot2)
library(ggtree)
library(knitr)
library(patchwork)
library(phangorn)
library(readr)
library(scales)
library(stringr)
library(tidyr)

data_dir <- params$data_dir

dataset_summary <- read_tsv(file.path(data_dir, "dataset_summary.tsv"), show_col_types = FALSE)
sample_qc <- read_tsv(file.path(data_dir, "sample_qc.tsv"), show_col_types = FALSE)
locus_summary <- read_tsv(file.path(data_dir, "locus_summary.tsv"), show_col_types = FALSE)
alignment_summary <- read_tsv(file.path(data_dir, "alignment_summary.tsv"), show_col_types = FALSE)
branch_metrics <- read_tsv(file.path(data_dir, "branch_metrics.tsv"), show_col_types = FALSE)
branch_alternatives <- read_tsv(file.path(data_dir, "branch_alternatives.tsv"), show_col_types = FALSE)
gene_heterogeneity <- read_tsv(file.path(data_dir, "gene_tree_heterogeneity.tsv"), show_col_types = FALSE)
topology_counts <- read_tsv(file.path(data_dir, "topology_counts.tsv"), show_col_types = FALSE)
species_tree <- read.tree(file.path(data_dir, "species_tree.report.tre"))

summary_row <- dataset_summary[1, ]

branch_metrics <- branch_metrics %>%
  mutate(
    species_tree_support_pct = 100 * as.numeric(species_tree_support),
    aster_local_posterior_pct = 100 * as.numeric(aster_local_posterior),
    aster_q1_pct = 100 * as.numeric(aster_q1),
    gcf = as.numeric(gcf),
    scfl = as.numeric(scfl)
  )

tree_plot_base <- ggtree(species_tree, size = 0.8) +
  geom_tiplab(size = 3.2) +
  geom_label2(
    aes(subset = !isTip & !is.na(label) & label != "", label = label),
    fill = "white",
    label.size = 0.15,
    size = 3
  ) +
  theme_tree2()

tree_order <- tree_plot_base$data %>%
  filter(!isTip, !is.na(label), label != "") %>%
  arrange(desc(y)) %>%
  pull(label)
```

## Takeaways

- Taxa analyzed: `r summary_row$sample_count`
- Candidate loci: `r summary_row$candidate_loci`; retained loci: `r summary_row$retained_loci`
- Sequence type: `r summary_row$sequence_type`
- Mean alignment length: `r round(as.numeric(summary_row$mean_alignment_length_sites), 1)` `r summary_row$sequence_length_unit`
- Complete-taxon gene trees: `r summary_row$complete_taxon_gene_tree_count`
- Weakest concordance branch: `r branch_metrics %>% arrange(gcf) %>% slice(1) %>% pull(label)`
- Weakest ASTER quartet branch: `r branch_metrics %>% arrange(aster_q1_pct) %>% slice(1) %>% pull(label)`

## Dataset Overview

This section summarizes taxon-level BUSCO QC, retained-locus occupancy,
alignment sizes, missingness, and filtering outcomes.

```{r}
#| label: dataset-overview-plots
#| fig-width: 12
#| fig-height: 9
busco_plot_data <- sample_qc %>%
  transmute(
    sample_id,
    single_copy = as.numeric(single_copy_percent),
    multi_copy = as.numeric(multi_copy_percent),
    fragmented = as.numeric(fragmented_percent),
    missing = as.numeric(missing_percent)
  ) %>%
  pivot_longer(-sample_id, names_to = "metric", values_to = "percent")

busco_plot <- ggplot(busco_plot_data, aes(x = reorder(sample_id, percent), y = percent, fill = metric)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Percent", fill = NULL, title = "BUSCO completeness by taxon") +
  theme_minimal(base_size = 11)

missingness_plot <- sample_qc %>%
  mutate(
    missing_pct = 100 * as.numeric(retained_missing_fraction),
    present_pct = 100 - missing_pct
  ) %>%
  select(sample_id, present_pct, missing_pct) %>%
  pivot_longer(-sample_id, names_to = "state", values_to = "percent") %>%
  ggplot(aes(x = reorder(sample_id, percent), y = percent, fill = state)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("present_pct" = "#3b7a57", "missing_pct" = "#c64e4e"), labels = c("Present", "Missing")) +
  labs(x = NULL, y = "Percent of retained loci", fill = NULL, title = "Retained-locus missingness by taxon") +
  theme_minimal(base_size = 11)

alignment_plot <- alignment_summary %>%
  mutate(alignment_length_sites = as.numeric(alignment_length_sites)) %>%
  ggplot(aes(x = alignment_length_sites)) +
  geom_histogram(binwidth = 25, fill = "#4c78a8", color = "white") +
  labs(x = paste("Alignment length (", summary_row$sequence_length_unit, ")", sep = ""), y = "Loci", title = "Alignment length distribution") +
  theme_minimal(base_size = 11)

filter_plot_data <- locus_summary %>%
  filter(decision != "retain") %>%
  separate_rows(failure_reasons, sep = ",") %>%
  filter(failure_reasons != "") %>%
  count(failure_reasons, sort = TRUE)

filter_plot <- ggplot(filter_plot_data, aes(x = reorder(failure_reasons, n), y = n)) +
  geom_col(fill = "#e17c05") +
  coord_flip() +
  labs(x = NULL, y = "Excluded loci", title = "Filtering outcomes") +
  theme_minimal(base_size = 11)

(busco_plot + missingness_plot) / (alignment_plot + filter_plot)
```

## Pipeline Overview

```{mermaid}
flowchart LR
  A[BUSCO-derived ortholog loci] --> B[MAFFT locus alignments]
  B --> C[IQ-TREE locus trees]
  C --> D[wASTRAL species tree]
  D --> E[IQ-TREE gCF/sCF on species tree]
  E --> F[Quarto HTML report]
```

## Main Species Tree

Internal branch labels (`B1`, `B2`, ...) provide the stable join key used in the
concordance and conflict panels below.

```{r}
#| label: main-species-tree
#| fig-width: 10
#| fig-height: 5
tree_plot_base
```

## Branch Confidence vs Concordance

```{r}
#| label: branch-confidence-vs-concordance
#| fig-width: 10
#| fig-height: 4.5
branch_metric_plot_data <- branch_metrics %>%
  select(label, species_tree_support_pct, aster_local_posterior_pct, aster_q1_pct, gcf, scfl) %>%
  pivot_longer(-label, names_to = "metric", values_to = "value") %>%
  mutate(
    label = factor(label, levels = tree_order),
    metric = recode(
      metric,
      species_tree_support_pct = "ASTER support",
      aster_local_posterior_pct = "ASTER local PP",
      aster_q1_pct = "ASTER quartet frequency",
      gcf = "gCF",
      scfl = "sCF"
    )
  )

ggplot(branch_metric_plot_data, aes(x = value, y = label, color = metric)) +
  geom_line(aes(group = label), color = "grey85", linewidth = 0.4) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Percent / score", y = "Species-tree branch", color = NULL) +
  theme_minimal(base_size = 11)
```

```{r}
#| label: branch-metrics-table
branch_metrics %>%
  transmute(
    Branch = label,
    Split = paste(split_left, "vs", split_right),
    `ASTER support` = round(species_tree_support_pct, 1),
    `ASTER q1` = round(aster_q1_pct, 1),
    gCF = round(gcf, 1),
    sCF = round(scfl, 1)
  ) %>%
  kable()
```

## Conflict and Alternative Resolutions

```{r}
#| label: conflict-plots
#| fig-width: 12
#| fig-height: 9
gcf_conflict_plot <- branch_alternatives %>%
  filter(source == "gcf") %>%
  mutate(
    value = as.numeric(value),
    label = factor(label, levels = tree_order),
    alternative_id = recode(alternative_id, main = "gCF", df1 = "gDF1", df2 = "gDF2", dfp = "gDFP")
  ) %>%
  ggplot(aes(x = label, y = value, fill = alternative_id)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Species-tree branch", y = "Percent", fill = NULL, title = "Gene-tree conflict (gCF)") +
  theme_minimal(base_size = 11)

scfl_conflict_plot <- branch_alternatives %>%
  filter(source == "scfl") %>%
  mutate(
    value = as.numeric(value),
    label = factor(label, levels = tree_order),
    alternative_id = recode(alternative_id, main = "sCF", df1 = "sDF1", df2 = "sDF2")
  ) %>%
  ggplot(aes(x = label, y = value, fill = alternative_id)) +
  geom_col() +
  scale_fill_brewer(palette = "Pastel1") +
  labs(x = "Species-tree branch", y = "Percent", fill = NULL, title = "Site conflict (sCF)") +
  theme_minimal(base_size = 11)

aster_conflict_plot <- branch_alternatives %>%
  filter(source == "aster") %>%
  mutate(
    value = 100 * as.numeric(value),
    label = factor(label, levels = tree_order),
    alternative_id = recode(alternative_id, t1 = "q1", t2 = "q2", t3 = "q3")
  ) %>%
  ggplot(aes(x = label, y = value, fill = alternative_id)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3") +
  labs(x = "Species-tree branch", y = "Quartet frequency (%)", fill = NULL, title = "ASTER quartet alternatives") +
  theme_minimal(base_size = 11)

(gcf_conflict_plot / scfl_conflict_plot) / aster_conflict_plot
```

```{r}
#| label: aster-alternative-table
branch_alternatives %>%
  filter(source == "aster") %>%
  mutate(value = round(100 * as.numeric(value), 1)) %>%
  select(Branch = label, Topology = alternative_id, `Quartet frequency (%)` = value, Split = split) %>%
  kable()
```

## Gene-tree Heterogeneity

This section focuses on complete-taxon gene trees, because branch-by-branch
topology comparisons are only directly comparable when the full taxon set is
present.

```{r}
#| label: heterogeneity-plots
#| fig-width: 12
#| fig-height: 8
rf_plot <- gene_heterogeneity %>%
  filter(is_complete_taxon == TRUE | is_complete_taxon == "true") %>%
  mutate(rf_distance_to_species_tree = as.numeric(rf_distance_to_species_tree)) %>%
  ggplot(aes(x = rf_distance_to_species_tree)) +
  geom_histogram(binwidth = 1, fill = "#4c78a8", color = "white") +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(x = "Robinson-Foulds distance to species tree", y = "Gene trees", title = "Gene-tree distance to the species tree") +
  theme_minimal(base_size = 11)

topology_plot <- topology_counts %>%
  slice_head(n = 10) %>%
  mutate(
    fraction = 100 * as.numeric(fraction),
    topology_id = factor(topology_id, levels = rev(topology_id))
  ) %>%
  ggplot(aes(x = topology_id, y = fraction, fill = matches_species_tree)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("true" = "#3b7a57", "false" = "#c64e4e"), labels = c("Species-tree topology", "Alternative topology")) +
  labs(x = NULL, y = "Complete-taxon gene trees (%)", fill = NULL, title = "Top gene-tree topologies") +
  theme_minimal(base_size = 11)

model_plot <- gene_heterogeneity %>%
  count(selected_model, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(selected_model, n), y = n)) +
  geom_col(fill = "#72b7b2") +
  coord_flip() +
  labs(x = NULL, y = "Loci", title = "Most common IQ-TREE models") +
  theme_minimal(base_size = 11)

support_plot <- gene_heterogeneity %>%
  count(support_values_present) %>%
  ggplot(aes(x = support_values_present, y = n, fill = support_values_present)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("true" = "#3b7a57", "false" = "#c64e4e")) +
  labs(x = "Support values present", y = "Gene trees", title = "Support coverage across gene trees") +
  theme_minimal(base_size = 11)

(rf_plot + topology_plot) / (model_plot + support_plot)
```

## Topological uncertainty

```{r}
#| label: topology-tests-load
#| include: false
topo_tests_available <- file.exists(file.path(data_dir, "au_test_results.tsv"))
if (topo_tests_available) {
  branch_quartet_support <- read_tsv(file.path(data_dir, "branch_quartet_support.tsv"), show_col_types = FALSE)
  contested_branches     <- read_tsv(file.path(data_dir, "contested_branches.tsv"), show_col_types = FALSE)
  au_test_results        <- read_tsv(file.path(data_dir, "au_test_results.tsv"), show_col_types = FALSE)
  candidate_manifest     <- read_tsv(file.path(data_dir, "candidate_trees_manifest.tsv"), show_col_types = FALSE)
}
```

```{r}
#| label: topology-tests-guard
#| echo: false
#| results: asis
if (!topo_tests_available) {
  cat("> **Note:** Topology test outputs are not present in this report run. ",
      "Set `run_topology_tests: true` in the config and re-run to enable this section.\n")
}
```

```{r}
#| label: quartet-posteriors
#| eval: !expr topo_tests_available
#| echo: false
branch_quartet_support %>%
  mutate(
    pp1 = round(as.numeric(pp1), 3),
    pp2 = round(as.numeric(pp2), 3),
    pp3 = round(as.numeric(pp3), 3),
    q1  = round(as.numeric(q1),  3),
    q2  = round(as.numeric(q2),  3),
    q3  = round(as.numeric(q3),  3),
    Contested = ifelse(as.logical(contested), "Yes", "")
  ) %>%
  select(Branch = branch_id, pp1, pp2, pp3, q1, q2, q3, Contested) %>%
  kable(caption = "Per-branch quartet posterior probabilities (pp1 = main topology, pp2/pp3 = NNI alternatives). Contested branches (pp1 < threshold) are flagged.")
```

```{r}
#| label: au-test-results
#| eval: !expr topo_tests_available
#| echo: false
au_test_results %>%
  mutate(
    logL  = round(as.numeric(logL), 2),
    deltaL = round(as.numeric(deltaL), 2),
    p_au  = round(as.numeric(p_au), 4),
    `Confidence set` = ifelse(as.logical(in_confidence_set), "Yes", "No")
  ) %>%
  select(
    `Tree` = tree_index, Description = description, Source = source,
    logL, deltaL, `p-AU` = p_au, `Confidence set`
  ) %>%
  kable(caption = "AU topology test results. Trees with p-AU ≥ 0.05 belong to the 95% confidence set and cannot be rejected.")
```

> **Interpretation guide**
>
> - **Quartet posteriors (pp1–pp3)**: Coalescent-aware per-branch support under the multispecies coalescent. A branch with pp1 > 0.95 is well resolved; pp1 ~ 0.5–0.7 indicates genuine topological ambiguity due to ILS or insufficient data.
> - **AU p-values**: Concatenation-based topology test. Trees with p-AU ≥ 0.05 cannot be rejected given the total sequence evidence.
> - **Agreement**: When a contested branch (low pp1) also has an alternative tree in the AU confidence set, both the coalescent and concatenation perspectives agree on the ambiguity — the strongest signal that the branch is genuinely unresolved.
> - **Disagreement**: Low pp1 but AU rejects the alternative may indicate systematic error (e.g., long-branch attraction) rather than ILS; high pp1 but AU keeps the alternative may indicate ILS that concatenation cannot model.

## Appendix

```{r}
#| label: topology-table
topology_counts %>%
  slice_head(n = 10) %>%
  mutate(fraction = round(100 * as.numeric(fraction), 1)) %>%
  select(Topology = topology_id, Count = count, `Percent (%)` = fraction, `Matches species tree` = matches_species_tree, Example = example_locus_id) %>%
  kable()
```
